The \project{unRAVE} catalog

### 2016-09-03 article/unrave.tex
unbalanced
Some text edits and incorporate Kordopatis comments sent by email

\begin{abstract}
The Milky Way may be our greatest laboratory for understanding galaxy formation.
This premise relies on the ability to precisely determine the astrometry and
chemistry for many individual stars, and to use those data to make inferences
about the Galaxy's formation and evolution. On 14 September 2016 the \gaia\ 
mission will release precise astrometry for $\approx2\times10^6$ stars.  
However, a comprehensive catalog of detailed chemical abundances to complement
those data is currently lacking.  The RAdial Velocity Experiment (\rave) survey 
has acquired spectra for 292,036 stars in common with the first \gaia\ data 
release, constituting the largest overlap of any stellar spectroscopic sample.  
Here we perform an independent analysis of \rave\ spectra using an implementation
of \thecannon\ that uses prior probabilities on atomic line formation and stellar
astrophysical parameters.  Our approach yields precise stellar labels at lower 
S/N ratios than most \emph{ab initio} approaches, allowing us to deliver improved
radial velocities, effective temperature $\teff$, surface gravity $\logg$, and 
chemical abundances of up to seven elements (O, Mg, Al, Si, Ca, Fe, Ni) for 
\Nstars\ stars.  We validate our results internally from repeat visits, and 
externally against existing high-resolution spectroscopic surveys.  The typical
precision in chemical abundances is X.XX~dex.  We provide spectrophotometric 
distances based on our labels, improving upon the parallax distances available
for the fainter targets in the first \gaia\ data release. \stub{A one-line 
science result based on RAVE/Hipparcos overlap.}
\end{abstract}

\keywords{}

@@

The Milky Way is considered to be the best laboratory for understanding galaxy
formation and evolution.  This premise hinges on the ability to precisely measure 
the astrometry and chemistry for (many) individual stars, and to use those data 
to infer the structure, kinematics, and chemical enrichment of the Galaxy 
\citep[e.g.,][]{Schlaufman_2009,Deason_2011,Ness_2012,Ness_2013a,Ness_2013b,
Casey_2012,Casey_2013,Casey_2014a,Casey_2014b,Bovy_2016}.  However, these 
quantities are not known for even 1\% of stars in the Milky Way.  Stellar 
distances are famously imprecise \citep[e.g.,][]{van_Leeuwen_2007,Jofre_2015,
Madler_2016}, proper motions can be plagued by unquantified systematics from 
the first epoch observations \citep[e.g.,][]{Casey_Schlaufman_2015}, and 
stellar spectroscopists frequently report significantly different chemical 
abundance patterns for the same star \citep{Smiljanic_2014}.  The impact 
these issues have on scientific inferences cannot be understated.  Imperfect 
astrometry or chemistry limits understanding in a number of sub-fields in
astrophysics, including the properties of exoplanet host stars, the formation 
(and destruction) of stars and clusters, as well as the structure of the 
Galactic disk, to name a few.

These formed just some of the scientific drivers for the \gaia\ space telescope.
\gaia\ is primarily an astrometric mission, providing precise positions,
parallaxes and proper motions for more than $10^9$ stars in its final data
release in 2022.  While this is a sample size about three orders of magnitude 
larger than its predecessor \hipparcos, both astrometry and chemistry are 
required to fully characterize the formation and evolution of our Galaxy. 
\gaia\ will also provide radial velocities, stellar parameters and chemical 
abundances for a subset of brighter stars, but these measurements will not be 
available in the first few data releases. Until those abundances are available,
astronomers seeking to use chemistry and kinematic measurements are reliant on 
ground-based spectroscopic surveys in order to complement the available astrometry.

The first \project{Gaia} data release will include positions, proper motions, and 
parallaxes for approximately two million stars in the Tycho-2 \citep{Hog_2000} 
catalog \citep{Michalik_2015a,Michalik_2015b}.  After cross-matching all major 
stellar spectroscopic surveys\footnote{Specifically we cross-matched the Tycho-2
catalog against the \apogee\ \citep{Zasowski_2013}, \ges\
\citep{Gilmore_2012,Randich_2013}, \galah\ \citep{DeSilva_2015},
\lamost\ \citep{Cui_2012}, and \rave\ \citep{Steinmetz_2006} 
surveys}, we found the RAdial Velocity Experiment (hereafter \rave) survey to 
have the largest overlap with the first \gaia\ data release: 292,036 
stars.  We then used the \gaia\ universe model snapshot \citep{Robin_2012} to 
estimate the precision in parallax and proper motions that will be available in 
the first \gaia\ data release (DR1) for stars in those overlap samples. Comparing
the expected precision to what is currently available, we further found that the
\rave\ survey will benefit most from \gaia\ DR1.  The distances of 63\% of the 
\rave--\gaia\ DR1 overlap sample (182,862 stars) are expected to improve with 
the first \gaia\ data release, and 47\% of stars are likely to have better 
proper motions (137,211 stars).  This motivated us to examine what chemical 
abundances were available for \rave, and to evaluate whether we could enable new 
chemodynamic studies by contributing to this existing set of chemical abundances.

We briefly describe the \rave\ data in Section~\ref{sec:data}, before explaining
our methods in Section~\ref{sec:method}.  In Section~\ref{sec:validation}
we outline a number of validation experiments, including: internal sanity checks,
comparisons with literature samples, and investigations to ensure our results
are consistent with expectations from astrophysics.  We discuss these results in
Section~\ref{sec:discussion}, and in Section~\ref{sec:conclusion} we provide 
concluding remarks as well as a description of how to access our results electronically.

@@

\rave\ is a magnitude-limited stellar spectroscopic survey of the (nearby) Milky Way,
principally designed to measure radial velocities for up to $10^6$ stars.
Observations were conducted on the 1.2~m UK Schmidt telescope at the Australian 
Astronomical Observatory\footnote{Formerly the Anglo-Australian Observatory} from 
2003--2013.  A large 5.7~degree field-of-view and robotic fibre positioner 
made for very efficient observing;  spectra for up to 150 targets could be simultaneously
acquired.  When observations concluded in April 2013, a total of \Nspectra\ useful spectra
had been collected of \Nstars\ unique objects.

The target selection for \rave\ is based on the $I$-band apparent magnitude,
$9 < I < 12$, with a weak $J - K_s > 0.5$ cut near the disk and bulge \citep{Wonjo_2016}.  
The $I$ band was selected because it overlaps with the wavelength range that \rave\ 
operates in:  8410--8795~\AA.  This spectral region is very well-studied.  It is one 
of the key setups used for the ground-based high-resolution \ges\ survey
\citep{Gilmore_2012,Randich_2013}, and the \rave\ resolution and wavelength
coverage is comparable to the Radial Velocity Spectrometer on board the \gaia\
space telescope \citep{Recio-Blanco_2016}.  The region includes the \ion{Ca}{2} 
near infrared triplet lines, strong transitions that are dominated by pressure 
broadening, which are visible even in very low S/N data or metal-poor stars.  
Atomic transitions of light-, $\alpha$-, and Fe-peak elements are also present, 
allowing for detailed chemical abundance studies.

@@

We sought to construct a training set consisting of stars across the main-sequence,
sub-giant, and the red giant branch.  We required stars with precisely measured
effective temperature $\teff$, surface gravity $\logg$, and elemental abundances
of Fe, Mg, Si, Ti, Ca, Al, and Ni.  This proved to be difficult because the magnitude
range of \rave\ does not overlap substantially with high-resolution spectroscopic
surveys.  The fourth internal data release of the \ges\ survey includes 
giant and main-sequence stars, but only 142 overlap with \rave.  The thirteenth data 
release from \sdss\ \citep{sloan_dr13} includes labels for \apogee\ stars on the
giant branch and (uncalibrated values for) the main-sequence, but our tests indicated
that the main-sequence labels suffered from significant systematic effects.  A flat, 
then `up-turning' main-sequence is present, and the metallicity gradient trends in 
the opposite direction with respect to $\logg$ on the main-sequence (i.e., metal-poor
stars incorrectly sit above the isochrones in a classical Hertzsprung-Russell diagram)
If we consider lower-resolution studies as potential training sets, there are 2,369
stars that overlap with \lamost\ --- of which 2,213 have positive S/N ratios in the 
$g$-band (\texttt{snrg}).  However, the labels are expectedly less precise given the
lower resolution, no detailed abundance information is available for main-sequence 
stars\footnote{Abundance information is available for \lamost\ stars from \citet{Ho_2016},
but that sample contains only giant stars}, and similar systematic trends are apparent
in the lower main-sequence.  The \galah\ survey has a good overlap with \rave, but 
results from that sample are not yet precise enough for our purposes.

@@

\epic\ differs from the \project{Kepler} input catalog because \epic\ does not 
benefit from having narrow-band $DDO_{51}$ photometry in order to aid dwarf/giant 
classification.  Despite this limitation the labels in the \epic\ catalog have 
already been shown to be accurate and trustworthy \citep{Huber_2016}.  However, 
when the posteriors are wide (i.e., the quoted confidence intervals are large) 
due to limited information available, it is possible that a star has been 
mis-classified.  This is most prevalent for sub-giants, where \citet{Huber_2016} 
note that $\approx55-70$\% of sub-giants are misclassified as dwarfs.  The 
probability of misclassification is usually quantified in the uncertainties of 
each star; most sub-giants that could be dwarfs have large confidence intervals.  
Therefore, requiring low uncertainties will decrease the total sample size, but 
will remove most misclassifications.  The situation is far more favourable for 
dwarfs and giants.  Only $1-4$\% of giant stars are misclassified as dwarfs, and 
about 7\% of dwarfs are misclassified as giants.  To summarise, the \epic\ labels 
with narrow confidence intervals are usually of high fidelity, and any spurious 
misclassifications would be clearly distinguishable if spectra were available.

@@

Here we cross-match our results against literature samples to verify that
our labels are accurate.  Because this work is an independent analysis, our
first point of reference is against the official \rave\ data releases.  These
comparisons are shown in Figures \ref{fig:rave-dr4-comparison} and 
\ref{fig:rave-dr5-comparison}, where we show our labels with respect to the
results in the fourth and fifth \rave\ data releases, respectively.  In order
to provide a fair comparison, we only show stars that meet a number of quality
flags in both samples. 
\stub{What selections qualify as a fair comparison in both samples?}
There is very good agreement in $\teff$, with a bias
and RMS of just XX~K and XXX~K, respectively.  We find our $\logg$ values are
closest to the `calibrated' \rave\ surface gravities, which make use of 
asteroseismic information to correct (increase) the $\logg$ values along the
giant branch.  The calibrated \rave\ surface gravities also correct the 
location of the red clump (by $\sim0.5$~dex) to be in the same position in our
results.  However, while our $\logg$ values generally agree with the calibrated
\rave\ values along the giant branch, our $\logg$ values do differ slightly 
along the main-sequence.  This is principally because the main-sequence in 
this work tends to taper down towards higher $\logg$ values at cooler 
temperatures, whereas the comparison sample tend to have a slightly flatter
main-sequence.  This difference is not likely to have a very significant 
effect on the detailed abundance or spectrophotometric distance determinations
between these studies.

@@

\stub{Any differences in RAVE abundances? Corrections or caution required?}

@@

@@

@@

>>>>

\begin{abstract}
The Milky Way may be our greatest laboratory for understanding galaxy formation.
This premise relies on the ability to precisely determine the astrometry and
chemistry for many individual stars, and to use those data to make inferences
about the Galaxy's formation and evolution. On 14 September 2016 the \gaia\ 
mission will release precise astrometry for $\approx2\times10^6$ stars.  
However, a comprehensive catalog of detailed chemical abundances to complement
those data is currently lacking.  The RAdial Velocity Experiment (\rave) survey 
has acquired spectra for 292,036 stars in common with the first \gaia\ data 
release, constituting the largest overlap of any stellar spectroscopic sample.  
We have performed an independent analysis of \rave\ spectra using \thecannon,
and here we report effective temperature $\teff$, surface gravity $\logg$, 
as well as chemical abundances of up to seven elements (O, Mg, Al, Si, Ca, Fe, 
Ni) for \Nstars\ stars.  We validate our results internally from repeat visits, 
and externally against existing high-resolution spectroscopic surveys.  The 
typical precision is XXX~K in $\teff$, X.XX~dex in $\logg$, and X.XX~dex for
detailed abundances.  We provide spectrophotometric distances based on our
labels, improving upon the parallax-based distances available for distant 
giants in the first \gaia\ data release.
\stub{A one-line science result based on RAVE/Hipparcos overlap.}

\end{abstract}

\keywords{\vspace{5em}}

@@

The Milky Way is considered to be our best laboratory for understanding galaxy
formation and evolution.  This premise hinges on the ability to precisely measure 
the astrometry and chemistry for (many) individual stars, and to use those data 
to infer the structure, kinematics, and chemical enrichment of the Galaxy 
\citep[e.g.,][]{Schlaufman_2009,Deason_2011,Ness_2012,Ness_2013a,Ness_2013b,
Casey_2012,Casey_2013,Casey_2014a,Casey_2014b,Kordopatis_2015,Bovy_2016}.  
However, these quantities are not known for even 1\% of stars in the Milky Way.  Stellar distances are famously imprecise \citep[e.g.,][]{van_Leeuwen_2007,
Jofre_2015,Madler_2016}, proper motions can be plagued by unquantified systematics 
from the first epoch observations \citep[e.g.,][]{Casey_Schlaufman_2015}, and 
stellar spectroscopists frequently report significantly different chemical 
abundance patterns from the same spectrum \citep{Smiljanic_2014}.  The impact 
these issues have on scientific inferences cannot be understated.  Imperfect 
astrometry or chemistry limits understanding in a number of sub-fields in
astrophysics, including the properties of exoplanet host stars, the formation 
(and destruction) of stars and clusters, as well as the structure of the 
Galactic disk, to name a few.

These formed just some of the scientific drivers for the \gaia\ space telescope.
\gaia\ is primarily an astrometric mission, and will provide precise positions,
parallaxes and proper motions for more than $10^9$ stars in its final data
release in 2022.  While this is a sample size about three orders of magnitude 
larger than its predecessor \hipparcos, both astrometry and chemistry are 
required to fully characterize the formation and evolution of our Galaxy. 
\gaia\ will also provide radial velocities, stellar parameters and chemical 
abundances for a subset of brighter stars, but these measurements will not be 
available in the first few data releases. Until those abundances are available,
astronomers seeking to simultaneously use chemical and dynamical information are
reliant on ground-based spectroscopic surveys in order to complement the 
available \gaia\ astrometry.

The first \gaia\ data release will include positions, proper motions, and 
parallaxes for approximately two million stars in the Tycho-2 \citep{Hog_2000} 
catalog \citep{Michalik_2015a,Michalik_2015b}.  After cross-matching all major 
stellar spectroscopic surveys\footnote{Specifically we cross-matched the Tycho-2
catalog against the \apogee\ \citep{Zasowski_2013}, \ges\
\citep{Gilmore_2012,Randich_2013}, \galah\ \citep{DeSilva_2015},
\lamost\ \citep{Cui_2012}, and \rave\ \citep{Steinmetz_2006} 
surveys}, we found the RAdial Velocity Experiment (hereafter \rave) survey to 
have the largest overlap with the first \gaia\ data release: 292,036 
stars.  We then used the \gaia\ universe model snapshot \citep{Robin_2012} to 
estimate the precision in parallax and proper motions that will be available in 
the first \gaia\ data release (DR1) for stars in those overlap samples.  Comparing
the expected precision to what is currently available, we further found that the
\rave\ survey will benefit most from \gaia\ DR1.  The distances of 63\% of the 
\rave--\gaia\ DR1 overlap sample (182,862 stars) are expected to improve with 
the first \gaia\ data release, and 47\% of stars are likely to have better 
proper motions (137,211 stars).  Although the \gaia\ universe model assumes 
end-of-mission uncertainties --- and does not account for systematics in the first
data release --- this calculation still provides intuition for the relative 
improvement the first \gaia\ data release can make to ground-based surveys.  The 
expected improvements for \rave\ motivated us to examine what chemical abundances
were available from those data, and to evaluate whether we could enable new 
chemodynamic studies by contributing to the existing set of chemical abundances.

We briefly describe the \rave\ data in Section~\ref{sec:data}, before explaining
our methods in Section~\ref{sec:method}.  In Section~\ref{sec:validation}
we outline a number of validation experiments, including: internal sanity checks,
comparisons with literature samples, and investigations to ensure our results
are consistent with expectations from astrophysics.  We discuss the implications
of these comparisons and our results in Section~\ref{sec:discussion}.  We provide
concluding remarks in Section~\ref{sec:conclusion}, as well as a description for
how to access our results electronically.

@@

\rave\ is a magnitude-limited stellar spectroscopic survey of the (nearby) Milky Way,
principally designed to measure radial velocities for up to $10^6$ stars.
Observations were conducted on the 1.2~m UK Schmidt telescope at the Australian 
Astronomical Observatory\footnote{Formerly the Anglo-Australian Observatory} from 
2003--2013.  A large 5.7~degree field-of-view and robotic fibre positioner made for 
very efficient observing:  spectra for up to 150 targets could be simultaneously
acquired.  When observations concluded in April 2013, a total of \Nspectra\ useful 
spectra had been collected of \Nstars\ unique objects.

The target selection for \rave\ is based on the $I$-band apparent magnitude,
$9 < I < 12$, with a weak $J - K_s > 0.5$ cut near the disk and bulge \citep{Wonjo_2016}.  
The $I$ band was used for the target selection because it has a good overlap with the
wavelength range that \rave\ operates in:  8410--8795~\AA.  This spectral region 
is very well-studied.  It is one of the key setups used for the ground-based 
high-resolution \ges\ survey \citep{Gilmore_2012,Randich_2013}, and the 
\rave\ resolution and wavelength coverage is comparable to the Radial Velocity 
Spectrometer on board the \gaia\ space telescope \citep{Kordopatis_2011,Recio-Blanco_2016}.  
The region includes the \ion{Ca}{2} near infrared triplet lines --- strong 
transitions that are dominated by pressure broadening --- which are visible even 
in very low S/N data or metal-poor stars.  Atomic transitions of light-, $\alpha$-, 
and Fe-peak elements are also present, allowing for detailed chemical abundance studies.

@@

We sought to construct a training set consisting of stars across the main-sequence,
sub-giant, and the red giant branch.  We required stars with precisely measured
effective temperature $\teff$, surface gravity $\logg$, and elemental abundances
of Fe, Mg, Si, Ti, Ca, Al, and Ni.  This proved to be difficult because the magnitude
range of \rave\ does not overlap substantially with high-resolution spectroscopic
surveys.  The fourth internal data release of the \ges\ survey includes 
giant and main-sequence stars, but only 142 overlap with \rave.  The thirteenth data 
release from \sdss\ \citep{sloan_dr13} includes labels for \apogee\ stars on the
giant branch and (uncalibrated values for) the main-sequence, but our tests indicated
that the \apogee\ main-sequence labels suffered from significant systematic effects.  A flat, 
then `up-turning' main-sequence is present, and the metallicity gradient trends in 
the opposite direction with respect to $\logg$ on the main-sequence (i.e., metal-poor
stars incorrectly sit above the isochrones in a classical Hertzsprung-Russell diagram).
If we consider lower-resolution studies as potential training sets, there are 2,369
stars that overlap with \lamost\ --- of which 2,213 have positive S/N ratios in the 
$g$-band (\texttt{snrg}).  However, the labels are expectedly less precise given the
lower resolution, no detailed abundance information is available for main-sequence 
stars\footnote{Abundance information is available for \lamost\ stars from \citet{Ho_2016},
but that sample contains only giant stars}, and similar systematic trends are apparent
in the lower main-sequence.  The \galah\ survey has a good overlap with \rave, but 
results from that sample are not yet precise enough for our purposes.

@@

\epic\ differs from the \project{Kepler} input catalog because \epic\ does not 
benefit from having narrow-band $DDO_{51}$ photometry in order to aid dwarf/giant 
classification.  Despite this limitation the labels in the \epic\ catalog have 
already been shown to be accurate and trustworthy \citep{Huber_2016}.  However, 
when the posteriors are wide (i.e., the quoted confidence intervals are large) 
due to limited information available, it is possible that a star has been 
mis-classified.  This is most prevalent for sub-giants, where \citet{Huber_2016} 
note that $\approx55-70$\% of sub-giants are misclassified as dwarfs.  The 
probability of misclassification is usually quantified in the uncertainties of 
each star; most sub-giants that could be dwarfs have large confidence intervals.  
Therefore, requiring low uncertainties will decrease the total sample size, but 
in practice it removes most misclassifications.  The situation is far more favourable for 
dwarfs and giants.  Only $1-4$\% of giant stars are misclassified as dwarfs, and 
about 7\% of dwarfs are misclassified as giants.  To summarise, the \epic\ labels 
with narrow confidence intervals are usually of high fidelity, and any spurious 
misclassifications would be clearly distinguishable if spectra were available.

@@

Here we cross-match our results against literature samples to verify that
our labels are accurate.  Because this work is an independent analysis, our
first point of reference is against the official \rave\ data releases.  These
comparisons are shown in Figures \ref{fig:rave-dr4-comparison} and 
\ref{fig:rave-dr5-comparison}, where we show our labels with respect to the
results in the fourth and fifth \rave\ data releases, respectively.  In order
to provide a fair comparison, we only show stars that meet a number of quality
flags in both samples. 
\stub{What selections qualify as a fair comparison in both samples?}
There is very good agreement in $\teff$, with a bias
and RMS of just XX~K and XXX~K, respectively.  We find our $\logg$ values are
closest to the `calibrated' \rave\ surface gravities, which make use of 
asteroseismic information to correct (increase) the $\logg$ values along the
giant branch.  The calibrated \rave\ surface gravities also correct the 
location of the red clump (by $\sim0.5$~dex) to be in the same position in our
results.  However, while our $\logg$ values generally agree with the calibrated
\rave\ values along the giant branch, our $\logg$ values do differ slightly 
along the main-sequence.  This is principally because the main-sequence in 
this work tends to taper down towards higher $\logg$ values at cooler 
temperatures, whereas the comparison sample tend to have a slightly flatter
main-sequence.  This difference is not likely to have a very significant 
effect on the detailed abundance or spectrophotometric distance determinations
between these studies \citep{Binney_2014}.

@@

\stub{Any differences in RAVE abundances? Corrections required due to having an unbalanced training set? Any caution required?}

@@

\bibitem[Binney et al.(2014)]{Binney_2014} Binney, J., Burnett, B., Kordopatis, G., et al.\ 2014, \mnras, 437, 351

@@

\bibitem[Kordopatis et al.(2011)]{Kordopatis_2011} Kordopatis, G., Recio-Blanco, A., de Laverny, P., et al.\ 2011, \aap, 535, A106

@@

\bibitem[Kordopatis et al.(2015)]{Kordopatis_2015} Kordopatis, G., Binney, J., Gilmore, G., et al.\ 2015, \mnras, 447, 3526
